 

### 2021-05-08 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor. This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].Therefore, light reaching forest floors that is incident to canopy tops can range from approximately 1-2% in tropical forest floors  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate [@bartemucciOverstoryInfluencesLight2006], to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997]. Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute sunflecks to shaded forest regions that receive ~1/3 of their carbon during photosynthesis through these short light exposures [@waySunflecksTreesForests2012]. Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

@@

The strength of this buffering is greater under dense-canopy forests than open-canopy forests [*Defrenne et al. 2021*]. Therefore, dense forests maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests tend to maintain higher daytime maximum temperature in the understory that are closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].

@@

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] below-canopy compared to open forests and nearby open areas. This trend is greater in wetter conditions, in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a]. RH, in combination with temperature, determines vapor pressure deficit (VPD), an important metric expressing the driving force of water loss from a leaf.  Tropical forests during the wet season tend to have higher VPD, therefore greater evaporative demand, in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

The strength of buffering decreases with altitude, moving to open-forests  [[@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], increases with wetter than drier regions [**REF**] and proximity to water such as in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

@@

Leaves adapt to their biophysical environment to maximize carbon assimilation by maintaining optimum leaf temperature in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. As a result, leaves can exhibit 'limited homeothermy' where they are cooler than air at high temperature and warmer than air at low temperature or 'megathermy' where they are drastically warmer than $T_{air}$. In some cases, 'poikilothermy', where $T_{leaf}$ can be equalled to $T_{air}$ [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]where stomatal conductance ($g_s$)  largely influences $T_{leaf}$ by facilitating heat dissipation, and leaf water content affects the rate of leaf warming [@kochDiurnalPatternsLeaf1994a; @fauset_differences_2018]. With sufficient $g_s$ leaves can regulate their temperature even in biophysical environment [**REF**], but limitation to $g_s$, driven by irradiance, $T_{air}$, VPD or drought curbs heat dissipation, drastically increasing leaf temperature.

Incoming solar radiation is a strong driver of leaf temperature, relative to $T_{air}$. With maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, but $g_s$ limitation can cause solar radiation to substantially elevate $T_{leaf}$ greater than $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** In closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in the upper canopies than lower canopies, due to midday stomatal closure [Slot et al. 2019]. This is seen in tropical forests [@fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a], semi-deciduous tropical forest during the wet season [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].

Greater light input in exposed upper canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. Across forest types, temperate deciduous upper canopy $T_{leaf}$ - $T_{air}$ has shown to be greater than $T_{leaf}$ - $T_{air}$ of upper canopy tropical moist and wet [@mauTemperateTropicalForest2018a].

@@

While light based acclimation in leaf traits is well established, research on solely height-based leaf trait variation is sparse due to the difficulty of isolating increased tree height from light.However, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig. `r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Isolating height (>35m) under constant light conditions (photon flux density) canopy stomatal conductance decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Morphologically, leaf size [Kaare H. Jensen and Maciej A. Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have shown to decrease, while leaf mass per area (at greater than 20m) to increase  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

@@

Within-individuals going from buffered and shaded canopy positions to exposed positions, leaf traits vary on the basis of biophysical demands [@casasSunShadeLeaves2011] and therefore exhibit commonly observed patterns in leaf anatomical, structural and biochemical traits at an intra-canopy gradient, pointed by several studies [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. These observations contribute to [understanding] leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

Going from shaded to exposed canopy position, morphological adaptation to dissapate heat increases. Sun leaves are often thicker, smaller, hairier, with greater dissection--lobes and serration--[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with more vertical leaf angles than shade leaves [@niinemetsAdjustmentFoliageStructure1998]. All, collectively regulate $T_{leaf}$-photosynthetic balance. On the other hand, because shaded leaves have greater effective leaf width [@fauset_differences_2018], boundary layer thickness, stomatal resistance and therefore longer thermal time constant, they exhibit lower convective cooling [@leighInfluenceLeafSize2017; Wright et al. 2019]. Thus, shade leaves remain cooler in buffered conditions but can become substanially warmer under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]

@@

Nitrogen distribution along the vertical gradient optimizes to prevent photoinhition for sun leaves during direct light and shade leaves during sunflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018]. Therefore, during growing season, $N_{area}$, along with LMA, leaf soluble sugars, starch and non-structural carbons increase with exposure [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al. 2014]. $P_{area}$ follows a similar trend, but with increased growth at higher light P uptake in some cases cannot keep up and thus shows variable results [@poorterMetaanalysisPlantResponses2019a]

@@

Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], LMA, photosynthetic capcity per area, and stomatal conductance increases with height compared to saplings, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Across  multi-species canopy and understory pasticity to light and biophysical condition differs along with inherent traits such as leaf shape, size, $gs$ levels and therefore influences $T_{leaf$. As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by stomatal conductance; and therefore their ability to photosynthesize as well. During the day, max transpiration rate and stomtal conductance increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004]. Thus, stomatal limitation is greater and steeper in exposed canopies than shaded leaves [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020].

@@

Leaf temperature beyond photosynthetic optimum temperature declines photosynthesis [**REF**]. Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

Photosynthesis will also be affected by heat stress, when photoinhibition reduces photosynthesis either as a result of photoprotective; non-photochemical quenching or related to irreversible leaf damage leading to leaf necrosis. At increasing temperatures, photosynthetic system II [PSII] is particularly vulnerable. Damage to PSII  can lead to decreased electron transport rates and photosynthetic failure. Irreversible damage to PSII is done at 40-60C [Baker 2008; @feeley_thermal_2020]. Leaves in upper canopy tend to experience more heat stress and greater photoinhibition than lower canopy leaves in conditions of water stress or drought. [@niinemetsWorldwideAnalysisWithincanopy2015]

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020],as well in vertical species profile in Puerto Rico forests, no trend in the optimum temperature for net photosynthesis was found [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and no significant trend in temperate trees [@mau_temperate_2018].

>>>>

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along a vertical gradient in forests globally, amount of light decreases from canopy tops to the forest floor. This trend is mediated by foliage LAI and density, where the gradient is more prounced in broad-leaf forests than needle-leaf forests [Fig. `r fig_NEON_vertical`; @lowmanForestCanopies1995; @aussenacInteractionsForestStands2000a; @vonarxSpatiotemporalEffectsForest2012a; @poorterMetaanalysisPlantResponses2019b; Chazdon & Fetcher, 1984; Smith et al., 2019; Tymen et al., 2017].Light reaching forest floors that is incident to the canopy top can range from approximately 1-2% in tropical forests  [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], 4-5% in mixed-wood boreal and northern temperate forests[@bartemucciOverstoryInfluencesLight2006] to 17-34% in needle-leaf forests [@baldocchiSeasonalVariationEnergy1997]. Small canopy gaps and wind-induced movement of canopy leaves and branches also contribute sunflecks to shaded forest regions that receive ~1/3 of their carbon during photosynthesis through these short light exposures [@waySunflecksTreesForests2012]. Additionally, in heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

@@

The strength of this buffering is greater under dense-canopy than open-canopy forests [*Defrenne et al. 2021*]. Where dense forests maintain daytime cooler lower maximum  temperature, and nighttime warmer higher minimum in the understory, relatively more than pine and boreal forests, nearby clearings [@vonarxSpatiotemporalEffectsForest2012a, *Defrenne et al. 2021*; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. On the other hand, open-forests maintain higher daytime maximum temperature in the understory closer to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a; @martin_boundary_1999; @ramboCanopyMicroclimateResponse2009] and  sometimes warmer nighttime understory temperatures than above canopy, due to long wave infrared radiation from the soil [@vonarxSpatiotemporalEffectsForest2012a].

@@

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity [RH] below-canopy than open forests and nearby open areas. This trend is greater in wetter conditions, in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a]. RH, in combination with temperature, determines vapor pressure deficit (VPD), an important metric expressing the driving force of water loss from a leaf.  Tropical forests during the wet season tend to have higher VPD, therefore greater evaporative demand, in the upper canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b].

@@

With altitude, the strength of buffering decreases, moving to open-forests  [[@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a; @rajsnerovaMorphologicalBiochemicalPhysiological2015; @liangForestTypeHeight2019], increases in wetter than drier regions [**REF**] and with proximity to water, for eg. in riparian forests. [@davisMicroclimaticBufferingForests2019; @macekMaximumAirTemperature2019]

@@

Leaves adapt to their biophysical environment to maximize carbon assimilation, by maintaining optimum leaf temperature in a given circumstance [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020]. As a result, leaves can exhibit 'limited homeothermy' where they are cooler than air at high temperature and warmer than air at low temperature or 'megathermy' where they are drastically warmer than $T_{air}$. In some cases, 'poikilothermy', where $T_{leaf}$ can be equalled to $T_{air}$ [@cavaleriColdbloodedForestsWarming2020; @drakeNoEvidenceHomeostatic2020].

@@

Leaf thermoregulation is a tradeoff between water use strategy and optimum leaf temperature [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. With sufficient stomatal conductance ($g_s$) leaves can regulate their temperature, by dissapating heat, even in drastic biophysical environment [**REF**] but when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, VPD or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

Incoming solar radiation is a strong driver of $T_{leaf}$, relative to $T_{air}$. With maximum $g_s$, leaves can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, but $g_s$ limitation can cause solar radiation to substantially elevate $T_{leaf}$ greater than $T_{air}$ [@cavaleriColdbloodedForestsWarming2020].

@@

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.** In closed-canopy forests with increased buffering capacity, large $T_{leaf} - T_{air}$ increases with height where higher $T_{leaf}$ are found in upper than lower canopies, due to midday stomatal closure [Slot et al. 2019]. This is seen in tropical forests [@fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a], semi-deciduous tropical forest during the wet season [@rey-sanchez_spatial_2016], and as well in a temperate forest [@mauTemperateTropicalForest2018a].

Greater light input in exposed upper canopy leaves can sometimes increase $T_{leaf_{max}}$ beyond temperature optimum for photosynthesis ($T_{opt}$) and $T_{air}$ making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@doughtyAreTropicalForests2008; @mauTemperateTropicalForest2018a]. Additionally, the upper canopy of temperate-deciduous species can have greater $T_{leaf}$ - $T_{air}$ than $T_{leaf}$ - $T_{air}$ of upper canopy tropical moist and wet forest species [@mauTemperateTropicalForest2018a].

@@

While light-driven plasticity in leaf traits is well established, research on solely height-driven leaf trait variation is sparse due to the difficulty of isolating increased tree height from light. However, both theory and empirical evidence indicate that other factors--for example, height on a tree or exposure to evaporative demand (Fig. `r fig_NEON_vertical`)-- are important in shaping at least a subset of traits.

In particular, hydraulic constraints increase with height on a tree [REFS; *McDowell-isotopes book chapter*; @cobleLightAcclimationOptimizes2015; *Couvreur et al., 2018*] such that gas exchange is strongly constrained by at the tops of(~100 m. tall *Sequoia sempervirens*; *Ambrose et al. 2009, 2010*).
Isolating height (>35m) under constant light conditions, show that canopy stomatal conductance decreases along with increasing $VPD$  [@schaferEffectTreeHeight2000; @ambroseEffectsHeightTreetop2010]. Morphologically, leaf size [Kaare H. Jensen and Maciej A. Zwieniecki, 2013] and lobation [@kusiPlasticLeafMorphology2020] have shown to decrease, while leaf mass per area (at greater than 20m) to increase  [@cobleLightAcclimationOptimizes2015; @kenzoHeightrelatedChangesLeaf2015; @rijkersEffectTreeHeight2000a; @chinLeafAcclimationLight2017].

@@

Within-individuals going from buffered [shaded] canopy positions to exposed, leaf traits vary on the basis of biophysical demands [@casasSunShadeLeaves2011] and therefore exhibit commonly observed patterns in leaf anatomical, structural and biochemical traits at an intra-canopy gradient, pointed by several studies [@sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020]. These observations contribute to [understanding] leaf temperature distribution across the gradient [@fauset_differences_2018; @michaletzEnergeticCarbonEconomic2016] Table `r table_leaf_traits` summarizes relevant traits for thermal sensitivity.

@@

Going from shaded to exposed canopy position, morphological adaptation to dissapate heat increases. Sun leaves are often thicker, smaller, hairier, with greater dissection--lobes and serration--[@leighInfluenceLeafSize2017; @vogelConvectiveCoolingLow1970; @sackHowStrongIntracanopy2006], hydraulically efficient [Brodribb et al. 2010], with more vertical leaf angles than shade leaves [@niinemetsAdjustmentFoliageStructure1998]. All, collectively regulate $T_{leaf}$-photosynthetic balance. On the other hand, because shaded leaves have greater effective leaf width [@fauset_differences_2018], boundary layer thickness, stomatal resistance and therefore longer thermal time constant, they exhibit lower convective cooling [@leighInfluenceLeafSize2017; Wright et al. 2019]. Which leads to cooler $T_{leaf}$ under buffered conditions, and substanially warmer $T_{leaf}$ under greater light or $T_{air}$ [@leighInfluenceLeafSize2017]

@@

Nitrogen distribution along the vertical gradient optimizes to prevent photoinhition in sun leaves during direct light and shade leaves during sunflecks under available soil nitrogen conditions [@kitaoCanopyNitrogenDistribution2018]. Therefore, during growing season, $N_{area}$, along with LMA, leaf soluble sugars, starch and non-structural carbons increase with exposure [@cobleLightDrivesVertical2014a; @cobleHowVerticalPatterns2016; Weerasinghe et al. 2014].

@@

Across tree-size ontogeny of the same species, while fundamental adaptations to light and water are observed in canopy and understory trees similar to within-species gradient [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012], LMA, photosynthetic capcity per area, and stomatal conductance increase with height compared to saplings, under constant light conditions [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Across  multi-species canopy and understory, pasticity to light and biophysical condition differs along with inherent traits such as leaf shape, size, $gs$ levels and therefore influences $T_{leaf}$. As a result, some species might be more susceptible to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

While exposed sun leaves are equipped to thermoregulate more efficiently than shaded leaves, they are largely constrained by $gs$ [@sanchesDifferentialLeafTraits2010; @hernandezSimilarTemperatureDependence2020]; and therefore their ability to photosynthesize as well. During the day, max transpiration rate and $gs$ increases with height and is greater in the upper canopy compared to shaded canopies [@kosugiSeasonalFluctuationsTemperature2006; @sanchesDifferentialLeafTraits2010]. During midday, stomtal closure occurs when evaporative demand exceeds water transport capacity in exposed leaves thereby increasing $T_{leaf}$ [@sanchesDifferentialLeafTraits2010; @zwienieckiHydraulicLimitationsImposed2004].

@@

Leaf temperature beyond photosynthetic optimum temperature declines photosynthesis due it's peaked response to temperature. Photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; @slotSituTemperatureResponse2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure and increasing leaf temperature (e.g. @slotSituTemperatureResponse2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ). Heat stress at $40-60{^\circ}C$ can cause irreversible damage to photosystem [II] leading to leaf necrosis and eventually leaf death [Baker 2008; @feeley_thermal_2020]. Leaves in tropical upper canopies have shown to experience air temperatures well above $40-60{^\circ}C$ on a regular basis, if $T_{leaf}$ is not regulated through $gs$ or adapated, heat stress can loss of canopy leaves [**REF**]

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), as optimum temperature $T_{opt}$ for $J_{max}$ has shown to increase with irradiance of an upper canopy leaf compared to shaded lower canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similar results were found for $T_{opt}$ of $V_{max}$ and $J_{max}$ of tropical sun and shade leaves [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], and across a vertical profile in tropical forests [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018].
